Preferences for mercantilization: the role of intergenerational social mobility and beliefs in meritocracy in Chile

### 2025-06-27 advances/informe3.tex
Up informe 3

What is the legitimate extent of market inequality in the eyes of the
public? Since the early 1980s, many countries have experienced a
widespread retreat from universal welfare programs and a shift toward
the privatization and commodification of public goods, welfare policies,
and social services (\citeproc{ref-gingrich_making_2011}{Gingrich,
2011}; \citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin
America, as elsewhere, neoliberal reforms reshaped welfare-state
institutions by extending market logic into domains of social
reproduction that were traditionally governed by the state
(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). This transformation
reduced the role of public provision and increased the presence of
private actors in core social services
(\citeproc{ref-harvey_breve_2015}{Harvey, 2015}). Echoing Polanyi's
(\citeproc{ref-polanyi_great_1975}{1975}) insight that markets
constitute a distinct moral order, the institutional diffusion of market
rules has fostered a corresponding moral economy: a constellation of
norms and values concerning fair allocation, embedded in institutions
and shaping individual subjectivities
(\citeproc{ref-mau_inequality_2015}{Mau, 2015};
\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}). Within
this framework, a growing body of research examines the extent to which,
and the mechanisms by which, citizens consider it fair that the
allocation of services like health care, pensions, and education be
governed by market-based criteria---a phenomenon known as \emph{market
justice preferences} (\citeproc{ref-busemeyer_skills_2014}{Busemeyer,
2014}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
2024}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
Understanding these preferences is crucial, as they contribute to
legitimizing economic inequality by framing it as the fair result of
individual responsibility and limited state intervention
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

@@

Market justice preferences are shaped not only by objective
socioeconomic conditions but also by popular beliefs about inequality.
Among these, meritocracy is a key normative principle underpinning
market-based distributive preferences
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). It frames inequality as
inevitable but justifiable through effort and talent
(\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
\citeproc{ref-young_rise_1958}{Young, 1958}). Studies show that
individuals with stronger meritocratic beliefs tend to perceive less
inequality and legitimize it by attributing economic differences to
personal achievement (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
2023}; \citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
\citeproc{ref-wilson_role_2003}{Wilson, 2003}). In highly unequal
societies where access to services is largely governed by market logic,
such beliefs play a critical role in normalizing inequality. Recent
evidence from Chile shows that students who believe effort and talent
are rewarded in their country express stronger preferences for
market-based access to healthcare, pensions, and education
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).

@@

The legitimacy of market-based inequalities is closely tied to beliefs
about distributive justice grounded in market principles. Beyond
stratification structures and socioeconomic conditions, the study of
inequality also focuses on individuals' beliefs about the origins of
inequality, the normative frameworks that sustain those beliefs, the
mechanisms that shape them, and their implications for attitudes and
behavior (\citeproc{ref-kluegel_beliefs_1981}{Kluegel \& Smith, 1981}).
In this regard, empirical research on distributive justice focuses on
individuals' conceptions of how goods and rewards \emph{should} be
distributed in society
(\citeproc{ref-jasso_distributive_2016}{Guillermina Jasso et al.,
2016}). This line of inquiry allows for examining the extent to which
economic inequality is perceived as just, or in a certain sense,
legitimate (\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011}).
Most of the research on this field explores the legitimacy of wage
inequality, specially salary gaps between jobs at opposite ends of the
occupational hierarchy
(\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011};
\citeproc{ref-jasso_justice_1978}{Guillermina Jasso, 1978};
\citeproc{ref-jasso_gender_1999}{G. Jasso \& Wegener, 1999}). Recently,
others examines how individuals justify market-generated inequalities in
access to core social services such as healthcare, education, pensions
or social security. Here, legitimacy stems from the belief that access
to these goods should follow market-based criteria
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such views, these
services are treated as legitimate commodities; goods that can be
traded, priced, and evaluated through market logic
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).

@@

The empirical study of market justice preferences has employed various
strategies to capture how individuals assess inequalities arising from
market allocation. A common approach evaluates whether people find it
fair that access to essential social services---such as healthcare,
education, or pensions---depends on income. This builds on foundational
work by Kluegel and Smith
(\citeproc{ref-kluegel_legitimation_1999}{1999}), who examined normative
justifications for capitalist systems. Subsequent studies have extended
this logic to a wider set of welfare goods. For example, Immergut and
Schneider (\citeproc{ref-immergut_it_2020}{2020}) and von dem Knesebeck
et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016}) explore whether
respondents believe it is just that higher-income individuals receive
better healthcare. Similarly, Lee and Stacey
(\citeproc{ref-lee_fairness_2023}{2023}) apply this framework to
education. Complementing these efforts, more recent research has
introduced composite indicators to capture broader orientations toward
market-based allocation. Lindh (\citeproc{ref-lindh_public_2015}{2015}),
for instance, constructs an index averaging support for income-based
access to healthcare and education in comparative perspective, while
Castillo et al. (\citeproc{ref-castillo_socialization_2024}{2024})
propose a single-item measure covering health, education, and pensions
in Chile. These instruments seek to gauge to what extent individuals
view market-generated inequalities as legitimate. Taken together, they
capture two core dimensions of market distribution: the role of economic
resources as a key determinant of outcomes, and the framing of services
as tradable commodities that can be bought and sold according to ability
to pay (\citeproc{ref-lindh_public_2015}{2015}).

Comparative empirical research has identified several individual-level
factors that influence support for market justice. Individuals in more
advantaged socioeconomic positions---those with higher income,
education, and occupational status---are consistently more likely to
endorse market-based distributive principles
(\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). For example,
Lindh (\citeproc{ref-lindh_public_2015}{2015}) finds that individuals
from the service class are more likely to support market-based access to
healthcare and education than skilled and unskilled workers across 17
relatively affluent countries. In a comparative analysis, Svallfors
(\citeproc{ref-svallfors_political_2007}{2007}) observes that this
expected class pattern appears clearly only in Sweden, where support for
private education and healthcare varies systematically by class.
Busemeyer (\citeproc{ref-busemeyer_skills_2014}{2014}) similarly shows
that support for private education is stronger among high-income groups,
while Immergut and Schneider (\citeproc{ref-immergut_it_2020}{2020}) and
von dem Knesebeck et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016})
report comparable findings for healthcare, suggesting that wealthier
individuals perceive private provision as a means of maintaining
relative advantage. In Chile, Otero and Mendoza
(\citeproc{ref-otero_power_2024}{2024}) show that individuals with
higher income and university education express stronger support for
market allocation in healthcare, education, and pensions. Beliefs about
inequality and political orientation also matter. In Chile, Castillo et
al. (\citeproc{ref-castillo_socialization_2024}{2024}) show that
individuals who have strong meritocratic beliefs are more likely to
support market-based distribution, while Lee and Stacy
(\citeproc{ref-lee_fairness_2023}{2023}) in Australia suggest that these
preferences are also greater as people lean toward economic
conservatism. In this sense, market justice preferences are shaped by
the interplay between structural position and normative reasoning.

@@

These dimensions of meritocracy reflect how respondents perceive
society's distributive logic, regardless of whether they endorse
meritocratic principles. This distinction aligns with recent findings
indicating that individuals distinguish between how merit is perceived
in society and how it should ideally operate, which in turn shapes their
preferences for redistribution and justice
(\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
2025}). Meritocratic beliefs serve as symbolic justifications for
unequal outcomes, particularly when access is stratified by income or
social opportunity. Prior studies have shown that individuals who
perceive higher levels of meritocracy tend to express stronger support
for unequal distributions that reflect market outcomes
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-castillo_meritocracia_2019}{Castillo et al., 2019}).

@@

@@

This study draws on data from the Chilean Longitudinal Social Survey
(ELSOC) of the Center for Social Conflict and Cohesion Studies (COES).
The survey is an annual panel collected from 2016 to 2023, comprising
two independent samples (original and refreshment), featuring permanent
and rotating questionnaire modules. It evaluates how individuals think,
feel, and behave regarding conflict and social cohesion in Chile. The
sampling design is complex, probabilistic, clustered, multistage, and
stratified by city size, targeting men and women aged 18--75 who
habitually reside in private dwellings in urban areas spanning 40 cities
(92 municipalities, 13 regions). This analysis focuses on 2023, the most
recent wave, which includes 1,737 respondents in the original sample and
989 in the refreshment sample. Further details on sampling, attrition,
and weighting can be found at https://coes.cl/encuesta-panel/, and the
dataset is publicly available at
https://dataverse.harvard.edu/dataverse/elsoc.

@@

@@

@@

@@

\bibitem[\citeproctext]{ref-davis_principles_2001}
Davis, K., \& Moore, W. E. (2001). Some {Principles} of
{Stratification}. In \emph{Social {Stratification}, {Class}, {Race}, and
{Gender} in {Sociological Perspective}, {Second Edition}} (2nd ed.).
Routledge.

@@

\bibitem[\citeproctext]{ref-espinoza_estratificacion_2013}
Espinoza, V., Barozet, E., \& Méndez, M. L. (2013). {Estratificaci{ó}n y
movilidad social bajo un}.

@@

\bibitem[\citeproctext]{ref-harvey_breve_2015}
Harvey, D. (2015). \emph{{Breve historia del neoliberalismo}}. Madrid
(Espa{ñ}a): Ediciones Akal.

@@

\bibitem[\citeproctext]{ref-jasso_justice_1978}
Jasso, Guillermina. (1978). On the {Justice} of {Earnings}: {A New
Specification} of the {Justice Evaluation Function}. \emph{American
Journal of Sociology}, \emph{83}(6), 1398--1419.
\url{https://doi.org/10.1086/226706}

\bibitem[\citeproctext]{ref-jasso_distributive_2016}
Jasso, Guillermina, Törnblom, K. Y., \& Sabbagh, C. (2016). Distributive
{Justice}. In C. Sabbagh \& M. Schmitt (Eds.), \emph{Handbook of {Social
Justice Theory} and {Research}} (pp. 201--218). New York, NY: Springer.
\url{https://doi.org/10.1007/978-1-4939-3216-0_11}

@@

@@

@@

@@

@@

>>>>

What is the legitimate extent of market inequality in the eyes of the
public? Since the early 1980s, many countries have experienced a
widespread retreat from universal welfare programs and a shift toward
the privatization and commodification of public goods, welfare policies,
and social services (\citeproc{ref-gingrich_making_2011}{Gingrich,
2011}; \citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin
America, as elsewhere, neoliberal reforms reshaped welfare-state
institutions by extending market logic into domains of social
reproduction that were traditionally governed by the state
(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). This transformation
reduced the role of public provision and increased the presence of
private actors in core social services
(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). Echoing Polanyi's
(\citeproc{ref-polanyi_great_1975}{1975}) insight that markets
constitute a distinct moral order, the institutional diffusion of market
rules has fostered a corresponding moral economy: a constellation of
norms and values concerning fair allocation, embedded in institutions
and shaping individual subjectivities
(\citeproc{ref-mau_inequality_2015}{Mau, 2015};
\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}). Within
this framework, a growing body of research examines the extent to which,
and the mechanisms by which, citizens consider it fair that the
allocation of services like health care, pensions, and education be
governed by market-based criteria---a phenomenon known as \emph{market
justice preferences} (\citeproc{ref-busemeyer_skills_2014}{Busemeyer,
2014}; \citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et
al., 2025}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
2024}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
Understanding these preferences is crucial, as they contribute to
legitimizing economic inequality by framing it as the fair result of
individual responsibility and limited state intervention
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

@@

Market justice preferences are shaped not only by objective
socioeconomic conditions but also by popular beliefs about inequality.
Among these, meritocracy is a key normative principle underpinning
market-based distributive preferences
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Studies show that
individuals with stronger meritocratic beliefs tend to perceive less
inequality and legitimize it by attributing economic differences to
personal achievement (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
2023}; \citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
\citeproc{ref-wilson_role_2003}{Wilson, 2003}). In highly unequal
societies where access to services is largely governed by market logic,
such beliefs play a critical role in normalizing inequality. Recent
evidence from Chile shows that students who believe effort and talent
are rewarded in their country express stronger preferences for
market-based access to healthcare, pensions, and education
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).

@@

The legitimacy of market-based inequalities is closely tied to beliefs
about distributive justice grounded in market principles. Beyond
stratification structures and socioeconomic conditions, the study of
inequality also focuses on individuals' beliefs about the origins of
inequality, the normative frameworks that sustain those beliefs, the
mechanisms that shape them, and their implications for attitudes and
behavior (\citeproc{ref-kluegel_beliefs_1981}{Kluegel \& Smith, 1981}).
In this regard, empirical research on distributive justice focuses on
individuals' conceptions of how goods and rewards \emph{should} be
distributed in society (\citeproc{ref-jasso_distributive_2016}{Jasso et
al., 2016}). This line of inquiry allows for examining the extent to
which economic inequality is perceived as just, or in a certain sense,
legitimate (\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011}).
Most of the research on this field explores the legitimacy of wage
inequality, specially salary gaps between jobs at opposite ends of the
occupational hierarchy
(\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011};
\citeproc{ref-jasso_justice_1978}{Jasso, 1978};
\citeproc{ref-jasso_gender_1999}{Jasso \& Wegener, 1999}). Recently,
others examines how individuals justify market-generated inequalities in
access to core social services such as healthcare, education, pensions
or social security. Here, legitimacy stems from the belief that access
to these goods should follow market-based criteria
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such views, these
services are treated as legitimate commodities; goods that can be
traded, priced, and evaluated through market logic
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).

@@

The empirical study of market justice preferences has employed various
strategies to capture how individuals assess inequalities arising from
market allocation. A common approach evaluates whether people find it
fair that access to essential social services---such as healthcare,
education, or pensions---depends on income. This builds on foundational
work by Kluegel and Smith
(\citeproc{ref-kluegel_legitimation_1999}{1999}), who examined normative
justifications for capitalist systems. Subsequent studies have extended
this logic to a wider set of welfare goods. For example, Immergut and
Schneider (\citeproc{ref-immergut_it_2020}{2020}) and von dem Knesebeck
et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016}) explore whether
respondents believe it is just that higher-income individuals receive
better healthcare. Similarly, Lee and Stacey
(\citeproc{ref-lee_fairness_2023}{2023}) and Castillo, Iturra, et al.
(\citeproc{ref-castillo_changes_2025}{2025}) apply this framework to
education. Complementing these efforts, more recent research has
introduced composite indicators to capture broader orientations toward
market-based allocation. Lindh (\citeproc{ref-lindh_public_2015}{2015}),
for instance, constructs an index averaging support for income-based
access to healthcare and education in comparative perspective, while
Castillo et al. (\citeproc{ref-castillo_socialization_2024}{2024})
propose a single-item measure covering health, education, and pensions
in Chile. These instruments seek to gauge to what extent individuals
view market-generated inequalities as legitimate. Taken together, they
capture two core dimensions of market distribution: the role of economic
resources as a key determinant of outcomes, and the framing of services
as tradable commodities that can be bought and sold according to ability
to pay (\citeproc{ref-lindh_public_2015}{Lindh, 2015}).

Comparative empirical research has identified several individual-level
factors that influence support for market justice. Individuals in more
advantaged socioeconomic positions---those with higher income,
education, and occupational status---are consistently more likely to
endorse market-based distributive principles
(\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). For example,
Lindh (\citeproc{ref-lindh_public_2015}{2015}) finds that individuals
from the service class are more likely to support market-based access to
healthcare and education than skilled and unskilled workers across 17
relatively affluent countries. In a comparative analysis, Svallfors
(\citeproc{ref-svallfors_political_2007}{2007}) observes that this
expected class pattern appears clearly only in Sweden, where support for
private education and healthcare varies systematically by class.
Busemeyer (\citeproc{ref-busemeyer_skills_2014}{2014}) similarly shows
that support for private education is stronger among high-income groups,
while Immergut and Schneider (\citeproc{ref-immergut_it_2020}{2020}) and
von dem Knesebeck et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016})
report comparable findings for healthcare, suggesting that wealthier
individuals perceive private provision as a means of maintaining
relative advantage. In Chile, Otero and Mendoza
(\citeproc{ref-otero_power_2024}{2024}) show that individuals with
higher income and university education express stronger support for
market allocation in healthcare, education, and pensions. Beliefs about
inequality and political orientation also matter. Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) and Castillo,
Laffert, et al. (\citeproc{ref-castillo_perceptions_2025}{2025}) show
that individuals who have strong meritocratic beliefs are more likely to
support market-based distribution in Chile, while Lee and Stacy
(\citeproc{ref-lee_fairness_2023}{2023}) in Australia suggest that these
preferences are also greater as people lean toward economic
conservatism. In this sense, market justice preferences are shaped by
the interplay between structural position and normative reasoning.

@@

These dimensions of meritocracy reflect how respondents perceive
society's distributive logic, regardless of whether they endorse
meritocratic principles. This distinction aligns with recent findings
indicating that individuals distinguish between how merit is perceived
in society and how it should ideally operate, which in turn shapes their
preferences for redistribution and justice
(\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
2025}). Meritocratic beliefs serve as symbolic justifications for
unequal outcomes, particularly when access is stratified by income or
social opportunity. Prior studies in Chile have shown that individuals
who perceive higher levels of meritocracy tend to express stronger
support for unequal distributions that reflect market outcomes in
healthcare, education and pensions
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
2024}).

@@

Chile offers a compelling case for examining individual preferences for
market justice in the access to core social services. Since the early
1990s, the country experienced a long period of sustained economic
growth, with per capita GDP increasing by nearly 4\% annually until the
mid-2010s (\citeproc{ref-llorca-jana_historia_2021}{Llorca-Jaña \&
Miller, 2021}). This growth brought notable improvements in living
standards, including a strong decline in income poverty---from 38.6\% in
1990 to 6.5\% in 2022 (\citeproc{ref-mideplan_evolucion_2017}{MIDEPLAN,
2017}). However, this success story has been tempered by growing
concerns over the unequal distribution of its benefits. In recent years,
economic growth has slowed---exacerbated by the global impact of the
COVID-19 pandemic---and structural inequalities have become more visible
(\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}). Despite its
economic achievements, Chile remains one of the most unequal countries
in Latin America and the OCDE. The poorest 50\% of the population
captures just 10\% of total income and holds negative net wealth, while
the wealthiest 1\% receives nearly 27\% of all income and controls
almost half (49.6\%) of the country's wealth
(\citeproc{ref-chancel_world_2022}{Chancel et al., 2022}), a figure that
has seen little change since the 1990s
(\citeproc{ref-flores_top_2020}{Flores et al., 2020}). These stark
disparities extend beyond income, reflecting deep inequalities in access
to essential social services (\citeproc{ref-pnud_desiguales_2017}{PNUD,
2017}).

A significant share of Chile's inequality stems from neoliberal reforms
that privatized and commodified key domains of social reproduction
(\citeproc{ref-arrizabalo_milagro_1995}{Arrizabalo, 1995};
\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). Introduced during the
military dictatorship (1973--1989) and expanded under democratic
governments, these reforms embedded market logic into public services
through concessions, subsidies, demand-side vouchers, and pro-private
regulatory frameworks (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). This
``crowded-out'' welfare model benefits higher-income groups, leaving
lower-income individuals to rely on limited public options. Scholars
argue that this neoliberal shift repurposed the state to create niches
of capitalist accumulation -from water to education-, giving rise to a
model of ``public-service capitalism'' heavily reliant on state funding
(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}). In healthcare, around
79\% of the population is covered by the public insurer FONASA, while
15\% are enrolled in private ISAPREs, which offer faster and
higher-quality services to wealthier groups
(\citeproc{ref-mideplan_estadisticas_2024}{MIDEPLAN, 2024}). The pension
system is based on individual capitalization, with mandatory
contributions managed by private administrators investing in the
financial market, currently involving 11 million contributors through
private fund administrators (AFPs), yet 27\% of the labor force remains
outside the system due to informality
(\citeproc{ref-superintendenciadepensiones_estadisticas_2024}{Superintendencia
de Pensiones, 2024}). Education is similarly stratified: only 30.6\% of
students attend fully public schools, while 54\% are in voucher-funded
private institutions and 9.3\% in fully private schools that largely
serve affluent families
(\citeproc{ref-ministeriodeeducacion_resumen_2023}{Ministerio de
Educación, 2023}). This arrangement has created a structurally segmented
system, where access and quality of services depend on ability to pay,
reinforced by state subsidies to private providers
(\citeproc{ref-pnud_desiguales_2017}{PNUD, 2017}).

Over recent decades, Chile has undergone a significant transformation in
its class structure, shaped by sustained economic growth, the expansion
of the service sector, the massification of education, and major changes
in the role of the state and social policy
(\citeproc{ref-ruiz_chilenos_2014}{Ruiz \& Boccardo, 2014}). These
changes contributed to the growth of the middle sectors, a phenomena
labelled as ``mesocratrization'' of society
(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013}).
However, this expansion did not reduce structural inequalities nor
facilitate upward mobility into elite positions
(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013};
\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}). While Chile shows high absolute mobility, most flows occur
between adjacent classes---particularly from working classes into lower
service-class positions---while access to upper classes remains limited,
indicating strong elite closure and reproduction
(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013};
\citeproc{ref-espinoza_movilidad_2014}{Espinoza \& Núñez, 2014};
\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}; \citeproc{ref-perez-ahumada_clases_2019}{Pérez-Ahumada, 2019};
\citeproc{ref-torche_unequal_2005}{Torche, 2005}). Though often
described as ``unequal but fluid'', recent evidence points to growing
rigidity in relative mobility, especially after the economic slowdown,
increased household debt, labor market precarity, and the COVID-19
crisis (\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}).
However, the majority of the population identifies themselves as members
of the middle class (\citeproc{ref-castillo_todos_2013}{Castillo et al.,
2013}). These dynamics have fueled a legitimacy crisis: meritocracy is
increasingly questioned, and many perceive that advancement depends more
on informal networks than on effort or earned credentials
(\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}). The
combination of constrained mobility, persistent inequality, and
concentrated opportunity has eroded expectations of upward mobility and
intensified social discontent (\citeproc{ref-pnud_desiguales_2017}{PNUD,
2017}).

Despite recurrent social unrest, Chile presents a paradoxical
coexistence between strong conflict over inequality and widespread
public legitimation of it. The October 2019 ``social outburst'', a
period of mass protests and severe political repression throughout the
country, crystallized discontent over the privatization and
commodification of social services, alongside a crisis of political
legitimacy (\citeproc{ref-somma_no_2021}{Somma et al., 2021}). A survey
during the protests identified pensions, healthcare, and education as
the top demands
(\citeproc{ref-nucleodesociologiacontingente_informe_2020}{Núcleo de
Sociología Contingente, 2020}), reflecting dissatisfaction with Chile's
stratified welfare regime. Yet, this unrest coexists with strong
meritocratic beliefs and high tolerance for income inequality. Castillo
(\citeproc{ref-castillo_legitimacy_2011}{2011}) shows that wage gaps
between occupations are consistently justified, and the larger the
perceived gap, the greater its legitimation, an effect amplified by
Chile's structural inequality. Moreover, perceptions of unfairness due
to non-meritocratic factors often reinforce belief in effort and talent
as legitimate bases for success
(\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}),
which are associated with lower perceived injustice
(\citeproc{ref-castillo_meritocracia_2019}{Castillo et al., 2019}).
Mac-Clure et al. (\citeproc{ref-mac-clure_justicia_2024}{Mac-Clure et
al., 2024}) find that low and lower-middle status individuals are less
likely than upper-status groups to view educational income differences
as unfair. Qualitative research further reveals the internalization of
the enterprising self (\citeproc{ref-mau_inequality_2015}{Mau, 2015}),
understood as a way in which the subjectivization process arises in line
with positioning the self in the market, generating its own market value
and subjecting it to competition. Canales et al.
(\citeproc{ref-canalesceron_sujeto_2021}{2021}) describes how families
navigate school choice as investment decisions, while Panes
(\citeproc{ref-panes_criticas_2020}{2020}) finds that many workers frame
pension contributions as individual investments aimed at maximizing
future returns. These representations suggest that market-based moral
orientations are deeply embedded in Chilean subjectivities, offering
fertile ground to study how support for market justice varies across
welfare domains.

Although still limited, recent research on market justice preferences in
Chile indicates that support for these beliefs has increased in recent
years across domains such as healthcare, pensions, and education, and
that both objective and subjective dimensions of inequality shape them.
Castillo et al. (\citeproc{ref-castillo_perceptions_2025}{2025}) show
that agreement with the notion that it is fair for higher-income
individuals to access better services was relatively low in 2016, but
rose significantly by 2023---especially in the pension domain. Otero and
Mendoza (\citeproc{ref-otero_power_2024}{2024}) provide evidence that
social class, measured using the EGP scheme, influences these
preferences: individuals in lower or subordinate class positions are
less supportive of market-based access to welfare services compared to
those in higher or privileged positions. Additionally, greater diversity
in social class networks is associated with lower support for market
justice. This is particularly relevant in the Chilean context, where one
of the main determinants of class-diverse networks is intergenerational
social mobility, which exposes individuals to relationships with people
from different class backgrounds (\citeproc{ref-otero_lives_2022}{Otero
et al., 2022}). Regarding subjective factors, Castillo et al.
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
2024}) find that individuals who endorse meritocratic beliefs---such as
the idea that effort and talent are rewarded in Chile---are more likely
to support market justice across welfare domains.

@@

This study draws on data from the Chilean Longitudinal Social Survey
(ELSOC) of the Center for Social Conflict and Cohesion Studies (COES).
The survey is an annual panel collected from 2016 to 2023, comprising
two independent samples (original and refreshment), featuring permanent
and rotating questionnaire modules. It evaluates how individuals think,
feel, and behave regarding conflict and social cohesion in Chile. The
sampling design is complex, probabilistic, clustered, multistage, and
stratified by city size, targeting men and women aged 18--75 who
habitually reside in private dwellings in urban areas spanning 40 cities
(92 municipalities, 13 regions). This analysis focuses on 2023, the most
recent wave, which includes 1,737 respondents in the original sample and
989 in the refreshment sample. Further details on sampling, attrition,
and weighting can be found at \url{https://coes.cl/encuesta-panel/}, and
the dataset is publicly available at
\url{https://dataverse.harvard.edu/dataverse/elsoc}.

@@

\bibitem[\citeproctext]{ref-arrizabalo_milagro_1995}
Arrizabalo, X. (1995). \emph{{Milagro o quimera: la econom{í}a chilena
durante la dictadura}}. Libros de la Catarata.

\bibitem[\citeproctext]{ref-barozet_clases_2021}
Barozet, E., Contreras, D., Espinoza, V., Gayo, M., \& Méndez, M. L.
(2021). \emph{{Clases medias en tiempos de crisis: vulnerabilidad
persistente, desaf{í}os para la cohesi{ó}n y un nuevo pacto social en
Chile}}. Santiago: Comisi{ó}n Econ{ó}mica para Am{é}rica Latina y el
Caribe (CEPAL).

@@

\bibitem[\citeproctext]{ref-castillo_changes_2025}
Castillo, J. C., Iturra, J., \& Carrasco, K. (2025). Changes in the
{Justification} of {Educational Inequalities}: {The Role} of
{Perceptions} of {Inequality} and {Meritocracy During} the {COVID
Pandemic}. \emph{Social Justice Research}.
\url{https://doi.org/10.1007/s11211-025-00458-0}

@@

\bibitem[\citeproctext]{ref-castillo_perceptions_2025}
Castillo, J. C., Laffert, A., Carrasco, K., \& Iturra, J. (2025).
Perceptions of {Inequality} and {Meritocracy}: {Their Interplay} in
{Shaping Preferences} for {Market Justice} in {Chile} (2016-2023).
\emph{Under Review at Frontiers in Sociology}.

\bibitem[\citeproctext]{ref-castillo_todos_2013}
Castillo, J. C., Miranda, D., \& Cabib, I. M. (2013). Todos somos de
clase media: {Sobre} el estatus social subjetivo en {Chile}. \emph{Latin
American Research Review}, \emph{48}(1), 155--173.
\url{https://doi.org/10.1353/lar.2013.0006}

@@

@@

\bibitem[\citeproctext]{ref-espinoza_estratificacion_2013}
Espinoza, V., Barozet, E., \& Méndez, M. L. (2013). {Estratificaci{ó}n y
movilidad social bajo un modelo neoliberal: El caso de Chile}.
\emph{Lavboratorio}, (25).

\bibitem[\citeproctext]{ref-espinoza_movilidad_2014}
Espinoza, V., \& Núñez, J. (2014). Movilidad ocupacional en {Chile}
2001-2009. {\textquestiondown}{Desigualdad} de ingresos con igualdad de
oportunidades? \emph{Revista Internacional de Sociolog{í}a},
\emph{72}(1), 57--82. \url{https://doi.org/10.3989/ris.2011.11.08}

@@

@@

\bibitem[\citeproctext]{ref-jasso_justice_1978}
Jasso, G. (1978). On the {Justice} of {Earnings}: {A New Specification}
of the {Justice Evaluation Function}. \emph{American Journal of
Sociology}, \emph{83}(6), 1398--1419.
\url{https://doi.org/10.1086/226706}

\bibitem[\citeproctext]{ref-jasso_distributive_2016}
Jasso, G., Törnblom, K. Y., \& Sabbagh, C. (2016). Distributive
{Justice}. In C. Sabbagh \& M. Schmitt (Eds.), \emph{Handbook of {Social
Justice Theory} and {Research}} (pp. 201--218). New York, NY: Springer.
\url{https://doi.org/10.1007/978-1-4939-3216-0_11}

@@

\bibitem[\citeproctext]{ref-llorca-jana_historia_2021}
Llorca-Jaña, M., \& Miller, R. M. D. (2021). \emph{{Historia econ{ó}mica
de Chile desde la independencia}}. Santiago de Chile: RIL editores.

@@

\bibitem[\citeproctext]{ref-mac-clure_justicia_2024}
Mac-Clure, O., Barozet, E., \& Franetovic, G. (2024). {Justicia
distributiva y posici{ó}n social subjetiva: {\textquestiondown}la
meritocracia justifica la desigualdad de ingresos?} \emph{Convergencia
Revista de Ciencias Sociales}, \emph{31}, 1.
\url{https://doi.org/10.29101/crcs.v31i0.22258}

@@

\bibitem[\citeproctext]{ref-mideplan_evolucion_2017}
MIDEPLAN. (2017). \emph{Evoluci{ó}n de la pobreza 1990-2017}. Ministerio
de Desarrollo Social y Familia - PNUD.

\bibitem[\citeproctext]{ref-mideplan_estadisticas_2024}
MIDEPLAN. (2024). \emph{Estad{í}sticas {Encuesta Casen} 2022, secci{ó}n
{Salud}}. Observatorio social.

@@

\bibitem[\citeproctext]{ref-ministeriodeeducacion_resumen_2023}
Ministerio de Educación. (2023). \emph{Resumen de matr{í}cula por
establecimiento educacional}. Datos abiertos Ministerio de Educaci{ó}n.

@@

\bibitem[\citeproctext]{ref-nucleodesociologiacontingente_informe_2020}
Núcleo de Sociología Contingente, {[}NUDESOC{]}. (2020). \emph{Informe
de resultados oficial {Encuesta Zona Cero}}. Santiago de Chile.
